\chapter{Reconciling the Two Semantics}
Chapters 2 and 3 have presented languages for defining the two distinct kinds of deterministic recurrences present in 
\cite{Karp}. In this chapter, we look at several attempts to connect these two recurrences. In particular, we want to find a 
way to write a recurrence with an initial condition as a recurrence of the form $T(x) = a(x) + T(m(x))$, as this is the kind
of recurrence for which Karp proves tail bounds. We will offer
several methods of translating from one expression language to the other, and investigate the validity of these 
approaches. Our goal here is to show that, for every expression of the form $\texttt{fix}(\lambda f.\lambda x.e)$ in our
 chapter 3 language, there are $a, m$ in our chapter 2 language such that 
 $\llbracket \texttt{rec}(a,m)\rrbracket_{sum} = \llbracket \texttt{fix}(\lambda f.\lambda x.e)\rrbracket_{fix}$. We use
 $\llbracket \cdot\rrbracket_{sum}$ and $\llbracket \cdot\rrbracket_{rec}$ to denote the semantics of chapters 2 and
 3, respectively. We will see that this will only be possible for $e$ of a certain form, but that this form covers all the 
 examples that Karp provides.

As a first attempt, we consider a simple recurrence with an initial condition, as described in chapter 3.
\begin{align*}
T(1) &= 0 \\
T(n) &= 1 + T(n/2 ) 
\end{align*}
As in the example in chapter 3, we assume that the domain of $T$ is positive powers of $2$.
We can write this recurrence as an expression \texttt{T} in our fixed-point language, where
\begin{align*}
\llbracket \texttt{T}\rrbracket_{fix}(n) = \log_2(n).
\end{align*} 

 In order to make this fit the definition of a recurrence described in 
chapter 2 --- that is, a recurrence of the form $T(n) = a(n) + T(m(n))$--- we may interpret $a$ as a 
piecewise function with two sub-functions: one for the general recurrence and one for the initial condition.
We write this as a single recurrence
\begin{align*}
T'(n) &= a(n) + T'(m(n)) \\
m(n) &= n/2, \ \forall n \\ 
a(n) &=
\begin{cases} 
0, \ n \leq 1 \\
1, \ n > 1
\end{cases}
\end{align*} 
This recurrence can then be expressed in our infinite sum language as 
\begin{align*}
\texttt{T} &= \texttt{rec}(a,m) \\
m &= \lambda x.x/\texttt{2} \\
a &= \lambda x.\ifelse{x\leq 1}{0}{1} \\
\llbracket \texttt{T} \rrbracket(n) &= \llbracket \texttt{rec}(a,m) \rrbracket(n) \\
&= \sum_{i=0}^{\infty}\llbracket a \rrbracket(\llbracket m\rrbracket^i(n)) \\
&= \sum_{i=0}^{\log(n)} 1 + \sum_{i=\log(n)+1}^{\infty} 0 \\
&= \log(n).
\end{align*}

However, this approach will not work in every case. Let us try to use the same process on the following recurrence
\begin{align*}
T(1) &= 1 \\
T(n) &= 2 + T(n/2) 
\end{align*}
We first write this as a single recurrence. See that, unlike the previous example, $a$ never maps to $0$.
\begin{align*}
T'(n) &= a(n) + T'(m(n)) \\
m(n) &= n/2, \ \forall n \\ 
a(n) &= 
\begin{cases}
1, \ n \leq 1 \\
2, \ n > 1
\end{cases}
\end{align*}
We then express this recurrence using our infinite sum language as
\begin{align*}
\texttt{T} &= \texttt{rec}(a,m) \\
m &= \lambda x.x/2 \\
a &=\lambda x.\ifelse{x\leq1}{1}{2} \\
\llbracket \texttt{T} \rrbracket(d) &= \llbracket \texttt{rec}(a,m) \rrbracket(d) \\
&= \sum_{i=0}^{\infty}\llbracket a \rrbracket(\llbracket m\rrbracket^i(d)) \\
&= \sum_{i=0}^{\log(n)} 2 + \sum_{i=\log(n)+1 }^{\infty} 1 \\
&= \infty
\end{align*}
Which, clearly, is not the same as the fixed-point semantics for $T$.

A second attempt takes a similar but slightly altered approach. Suppose that we interpret $a$ as a piecewise function with three sub-functions:
one for the general recurrence, one for the initial condition, and one for values smaller than the initial condition, which all
map to $0$. Considering the example that failed in the previous attempt, we would have
\begin{align*}
T'(n) &= a(n) + T'(m(n)) \\
m &= n/2, \forall n \\
a(n) &=
\begin{cases}
0, \ n < 1 \\
1, \ n = 1 \\
2, \ n > 1
\end{cases}
\end{align*}
We can express this recurrence using our infinite sum syntax as
\begin{align*}
\texttt{T} &= \texttt{rec}(a,m) \\
m &= \lambda x.x/\texttt{2} \\
a &= \lambda x.\ifelse{x <1}{0}{(\ifelse{x = 1}{1}{2})}.
\end{align*}
See, then, that
\begin{align*}
\llbracket a \rrbracket(n) &= 
 \begin{cases}
0, \ n < 1 \\
1, \ n = 1 \\
2, \ n > 1 
\end{cases} 
\end{align*}
\begin{align*}
\llbracket \texttt{T} \rrbracket(n) &= \llbracket \texttt{rec}(a,m)\rrbracket(n) \\
&= \sum_{i=0}^{\infty}\llbracket a \rrbracket(\llbracket m\rrbracket^i(n)) \\
&= \sum_{i=0}^{\log(n)} 2 + \sum_{i=\log(n)+1 }^{\log(n)+1} 1 + \sum_{i = \log(n) + 2}^{\infty} 0 \\
&= 2\log(n) + 1
\end{align*}

This approach matches our intuition for how an initial condition should work---that once you reach the initial condition,
you stop recursing. 

With this in mind, we can identify a general strategy for extracting a recurrence in our chapter 2 syntax (with
the $\texttt{rec}$ operator), given a recurrence in our chapter 3 syntax (with the $\texttt{fix}$ operator). 
Suppose we have a recurrence with an initial condition, given as follows:
\begin{align*}
T(b) &= c, \text{ where } b, c  \in \R \\
T(x) &= a(x) + T(m(x)).
\end{align*}
We write this recurrence in our chapter 3 syntax as 
\begin{align*}
\texttt{fix}(\lambda f.\lambda x. \ifelse{x = b}{c}{a(x) + f(m(x))}),
\end{align*}
an expression whose denotation will be a function $F$ such that
\begin{align*}
F(x) &=
\begin{cases}
c \text{ if } x = b \\
a(x) + c \text{ if } m(x) = b \\
a(x) + a(m(x)) + c \text{ if } m(m(x)) = b \\
\ldots \\
\sum_{i=1}^{n} a(m^{i-1}(x)) + c \text{ if } m^n(x) = b \\
\ldots
\end{cases}
\end{align*}

We may then write this recurrence in our chapter 2 syntax as $\texttt{rec}(a',m)$, where $m$ is the same function
as in the previous recurrence, and
\begin{align*}
a' = \lambda x.\lambda f. \ifelse{x < b}{0}{(\ifelse{x = b}{c}{a(x)})}.
\end{align*}
Then the denotation of $\texttt{rec}(a',m)$ is a function $F'$ such that for all $x$ in the domain of $F$, 
\begin{align*}
F'(x) &= \sum_{i=0}^{\infty} \llbracket a' \rrbracket(\llbracket m \rrbracket^i(x)) \\
&= \sum_{i=0}^{n-1} \llbracket a' \rrbracket( \llbracket m \rrbracket^i(x)) + \sum_{i=n}^n c + \sum_{i = n+1}^{\infty} 0 \\
&\text{ where } \llbracket m \rrbracket^n(x) = b \\
&= \sum_{i=1}^{n}\llbracket a' \rrbracket ( \llbracket m \rrbracket^{i-1}(x)) + c
\end{align*}
See, then, that $F' = F$. Therefore, this confirms that, given an expression $e$ of the specific form
$e = \ifelse{x = m}{c}{a(x) + f(m(x))}$, we can extract a function $a'$ such that 
\begin{align*}
\llbracket \texttt{fix}(\lambda f. \lambda x.e) \rrbracket_{fix} = \llbracket \texttt{rec}(a',m)\rrbracket_{sum}
\end{align*}





 